Self-organization of whisker barrels in simulation requires competition between thalamocortical axons

### 2019-11-19 paper/briefpaper/brief.tex
Housekeeping, creating longpaper and briefpaper folders.

% Whisker barrel generation paper equations, for pasting into Google
% docs draft.

% While writing, don't stop for errors
%\nonstopmode

\documentclass[a4paper,11pt]{article}
\usepackage{amsmath,amssymb,mathtools}
\usepackage{enumerate}
\usepackage{times}
\usepackage[usenames]{color}
\usepackage{parskip}
\usepackage{afterpage}
\usepackage{setspace}
\usepackage[modulo]{lineno}
%\usepackage{xltxtra}
%\usepackage{xunicode}
\usepackage{gensymb}

\usepackage{titlesec}
\titleformat*{\section}{\LARGE\bfseries}
\titleformat*{\subsection}{\Large\bfseries}
\titleformat*{\subsubsection}{\large\bfseries}
\titleformat*{\paragraph}{\large\bfseries}
\titleformat*{\subparagraph}{\large\bfseries}

\setcounter{tocdepth}{3}
\usepackage{graphicx}
\usepackage{comment}
\usepackage{textcomp}
\usepackage{natbib}

\usepackage{lmodern}

\usepackage[labelfont=bf,labelsep=period,justification=justified]{caption}

\doublespacing
\topmargin 0.0cm
\oddsidemargin 0.5cm
\evensidemargin 0.5cm
\textwidth 16cm
\textheight 21cm
\makeatletter
\makeatother
\date{}
\pagestyle{plain}

% Blue for comments
\newcommand{\cmnt}[1]{\textcolor{blue}{#1}}
% Code in sans serif
\newcommand{\code}[1]{\textsf{#1}}
% Divergence symbol (Del dot Del)
\newcommand{\dvrg}{\nabla\vcdot\nabla}
% Emphasis shortcut
\newcommand{\e}{\emph}
% Bold shortcut
\newcommand{\bol}{\textbf}
% Math bold
\newcommand{\mb}[1]{\mathbf{#1}}
\makeatletter
\newcommand*\vcdot{\mathpalette\vcdot@{.35}}
\newcommand*\vcdot@[2]{\mathbin{\vcenter{\hbox{\scalebox{#2}{$\m@th#1\bullet$}}}}}
\makeatother

%\bibliographystyle{plos2015}
\bibliographystyle{apalike}

\begin{document}
%\begin{center}
\begin{flushleft}
{\huge
\textbf{Self-organization of whisker barrels in simulation requires competition between thalamocortical axons}
 \vspace{5em}
}
%\end{center}

Sebastian~S.~James$^{1*}$, Stuart~P.~Wilson$^{1}$ \vspace{1em}\\
$^{1}$Department of Psychology, The University of Sheffield, Sheffield, United Kingdom.\vspace{1em}\\
$^{*}$Corresponding author email: Seb.James@Sheffield.ac.uk\vspace{1em}\\
Short title: Self-organising whisker barrels\vspace{5em}\\
\end{flushleft}

\newpage{}
%\raggedright

\section*{Abstract}

Understanding how genetic information and spontaneous pattern formation
together give rise to the whisker barrels is an important challenge for
developmental neurobiology. The barrel patterning that coalesces in layer IV
has been described formally as a Voronoi tesselation, which is indicative of
an underlying developmental process based on local competitive
interactions. If pattern formation relies only on such intrinsic
self-organising processes, how can the somatotopic order at the periphery be
maintained in the projections from brainstem to thalamus and from thalamus to
cortex? To investigate, we began by extending a computational model of
neocortical arealization by Karbowski \& Ermentrout (2004; J. Comp. Neurosci.)
to generate fields on a two-dimensional cortical sheet. Central to this model
is an assumption that axonal branching and synaptogenesis are coupled, and
that thalamic projections compete for cortical territory under the influence
of molecular signalling gradients. We found that these processes alone are
incapable of generating patterns of cortical fields that are more intricate
than the patterns already present in molecular guidance signals. However, a
full barrel-field pattern emerges under the guidance of as few as four
signalling gradients when the model is extended to introduce competition
between the branches of axons that originate from different thalamic
barreloids. We show that the key requirements for the emergence of realistic
(Voronoi tesselation) patterning are i) at each cortical location
thalamocortical projections compete for a limited number of available
connections, ii) at each location the axonal branching rate of each
thalamocortical projection is reduced by the total branching density of the
other projections, and iii) the density of axon branching for each
thalamocortical projection is conserved. We demonstrate how a purely
self-organising system can initialise a pattern which can then be faithfully
replicated in multiple neural structures. We show that this minimal model of
barrel self-organisation provides a parsimonious account of data from a range
of classic studies in which intrinsic and extrinsic factors have been
experimentally manipulated.

\section*{Introduction}

@@

Spatial patterns in neural connectivity provide clues about the constraints
under which brains evolve and develop. Whether or not patterning governs
information processing, models of the development of information processing
circuits can be validated in terms of their ability to account for biological
patterning (\citealp{purves_iterated_1992}; see also
\citealp{wilson_what_2015,bednar_cortical_2016}). One of the most distinctive patterns of
neural connectivity is found in the rodent barrel cortex
(\citealp{woolsey_pattern_1948,woolsey_structural_1970,welker_structure_1974}, see
\citealp{fox_barrel_2008}). Understanding how this particular pattern forms might shed
light on the interplay of genetic and extrinsic factors that shape cortical
circuits, and on a wealth of anatomical, physiological, and behavioural data
obtained using this model system.

@@

We first verified that the system given by Eqs.~\ref{eq:dc} \& \ref{eq:da}
described a two dimensional extension of the axon branching model in
\citet{karbowski_model_2004}. We showed that in the presence of three guidance
gradients varying only in one dimension (along the long axis of an elliptical
domain), five thalamocortical axon types, each interacting uniquely with the
three guidance gradients, would segregate into five separate regions. We
investigated a number of configurations of the guidance gradients, varying
their shape, $\rho(\mb{x})$, and number, $M$, and their effect on a varying
number of thalamocortical projections, $N$. In general, we were not able to
producen a complexity of structure in the resulting patterns of the
connections, $c(\mb{x},t\rightarrow\infty)$, which was not present in the
guidance patterns.

@@

To introduce competition between the \emph{branching} of thalamocortical axon
types, we introduced the new term $\chi$ given by
%
\begin{equation} \label{eq:branch_comp}
\chi(\mb{x}, t) = - \frac{\epsilon  a_i(\mb{x}, t)}{N-1} \sum_{i' \ne i}^{N} a_{i'}(\mb{x}, t).
\end{equation}
%
which leads to a reduction in the rate of increase of branching density for
axon type $i$ in regions where the branching density of \e{other} axon types
is high.

@@

With the addition of the competition term, $\chi$, it is necessary to
introduce a normalisation step, in order to prevent one TC type from
out-competing every other TC type. Normalisation represents the idea that for
a fixed number of afferent thalamocortical axons, the overall number of
branches must have some upper limit; given the finite resources available in
each axon for building new processes. We divisively normalize the branching
density, dividing it at each computational step by the mean branching density
per unit area; formalised as:
%
\begin{equation} \label{eq:norm}
a_i(\mathbf{x}, t) = A \; \frac {a'_i(\mathbf{x}, t)} {\int_\Omega a'_i(\mathbf{x}, t) d\sigma}
\end{equation}
%
where the prime symbols indicate the use of the intermediate value of $a_i$
after it has been computed according to Eqs.~\ref{eq:dc}--\ref{eq:branch_comp}
on the basis of the system state at time $t-{\delta}t$. $A$ is the area of the
domain. This normalises $a_i(\mathbf{x}, t)$ to 1, integrated over the domain. To
account for the possibility that there might be more axons afferent for TC
type $1$ cf.~TC type $2$, we normalise to the integral of $a_i$ at time 0, so
that Eq.~\ref{eq:norm} becomes:
%
\begin{equation} \label{eq:norm_time0}
\begin{split}
a_i(\mathbf{x}, t) & =
\frac{\int_\Omega  a_i(\mathbf{x}, 0) d\sigma}{A} \; A \; \frac {a'_i(\mathbf{x}, t)} {\int_\Omega
  a'_i(\mathbf{x}, t) d\sigma} \\
%
& = \int_\Omega  a_i(\mathbf{x}, 0) d\sigma \; \frac {a'_i(\mathbf{x}, t)} {\int_\Omega
  a'_i(\mathbf{x}, t) d\sigma}
%\\
%& = a_0 \frac {a'_i(\mathbf{x}, t)} {\int_\Omega
%  a'_i(\mathbf{x}, t) d\sigma}
\end{split}
\end{equation}
%
In the simulation, in which space is represented as $n_e$ discrete
hexagonal tiles, the equivalent expression is:
%
\begin{equation} \label{eq:norm_discrete}
  {a_{i,p}}(t) = \textstyle \sum_j^{n_e} a_{i,j}(0) \; \frac {a'_{i,p}(t)}
  {\textstyle \sum_j^{n_e} a'_{i,j}(t) }
\end{equation}
%
where $a_{i,j}(t)$ is the value of the variable $a_i$ in the $j^{\mathrm{th}}$
hexagonal element at time t.

We first investigated the effect of the competition by examining a system of
25 thalamocortical axon types simulated on a square domain, as in
Fig.~\ref{fig:sq}, to determine if a complex pattern could emerge. The hex
to hex distance was 0.02 mm and the square's side length was 1.16 mm (3919 hex
elements). The axons interacted with two orthogonal signalling gradients, one
of which increased from the top edge to the bottom edge, the other of which
had its expression increasing from the right edge to the left edge
[Fig.~\ref{fig:sq}(f)]. For each thalamocortical type, $i$, two
interaction parameters were set; $\gamma_{i,1}$ was the interaction with
guidance molecule 1 and $\gamma_{i,2}$ was the interaction with molecule
2. The branching of TC types with a negative interaction with molecule 2
($\gamma_{i,2}<0$) were pushed to the left; those with $\gamma_{i,2}>0$ were
pushed to the right. TC types that interacted negatively with molecule 1
($\gamma_{i,1}<0$) were pushed up; those with $\gamma_{i,1}>0$ were pushed
down.

\begin{figure*}
\begin{center}
\includegraphics[width=\textwidth]{./Fig_25N2M_square/Fig_25N2M_square.png}
\end{center}
%\vspace{-0em}
\caption{\textbf{(a)--(e)} Connection densities of 25 thalamocortical axons
  simulated on a square domain. \textbf{(a)} Connection density at simulation
  time t=0.005. The left image shows a map of the identity of the thalamocortical
  axon type with the highest absolute connection density. Hexagonal elements
  coloured the darkest blue have the highest connection density for TC type
  $i=1$; those coloured the darkest red have maximal connection density in TC
  type $i=25$. The right hand image of each pair shows the hexes for which the
  connection density passes 0.5 for each TC type, giving a contour map. At
  t=0, connection densities are randomly distributed. \textbf{(b)} t=0.18. The
  grid-like organisation of the TC connections is beginning to emerge.
  \textbf{(c)} t=0.25. The 5x5 grid is now clear in the left map. Branching
  competition and differential interaction with the two guidance gradients is
  causing the branching, and therefore the connection densities, to localise
  in 25 different locations. \textbf{(d)} t=0.81. Connection densities for
  each TC type have increased sufficiently for the contours to form. In (d),
  the mapping between colour and TC type $i$ is indicated. \textbf{(e)}
  t=1.77. Contours are well formed. In (e), the $\gamma$ values for each TC
  type, $i$, are shown. $\gamma_{i,1}$ gives the interaction with molecule `1'
  [red map in (f)]; $\gamma_{i,2}$ is the interaction with molecule
  `2'. \textbf{(f)} Expression maps for the two guidance molecules (`1' and
  `2') for this simulation. Colour saturation indicates expression intensity,
  which for each molecule varies linearly in one direction.}
\label{fig:sq}
\end{figure*}

Fig.~\ref{fig:sq}(a) shows the connection densities after the first 100
simulation time steps (t=0.005 s). Each colour uniquely identifies the TC
type, $i$ [the full mapping between index $i$ and the colour is shown in
  Fig.~\ref{fig:sq}(d)]. The left-hand map colours each hex according to
which TC type has the highest connection density. At this early stage of the
simulation, the colouring is somewhat random because the branching density was
initialised stochastically ($a_{i,j}(0) = 0.2 + U([0,0.2])$), which resulted
in a randomised pattern for ${c_i}^{max}$. The right-hand map of
Fig.~\ref{fig:sq}(a) colours those hexes which have $c_i \geq 0.5$ and an
adjoining hex for which $c_i < 0.5$ according to the mapping shown in
Fig.~\ref{fig:sq}(d)]. Due to the initial stochasticity, there are many
  such crossings of the threshold. The predominance of blue (for $i<10$)
  results from the order in which the hexes were rendered, with blue hexes
  being coloured last.  Fig.~\ref{fig:sq}(b) shows the simulation at
  t=0.14.
%
%
Fig.~\ref{fig:sq}(b) shows the connection density after 0.18 s of simulation
time (3600 steps of 5x10$^{-5}$ s). The initial randomness in the pattern has
disappearing, to be replaced by coalesced regions in which each TC type $i$ is
maximal.  There is a clear blue/red division from left to right. The threshold
plot indicates that the guidance molecules have pushed the branching towards
the corners of the square for the 4 TC types with the strongest interactions
with the guidance molecules; these exceed the threshold of 0.5. For the other
21 TC types, the connection densities are not yet sufficiently localised to
cross the threshold.
%
0.08 s of simulation time later, in Fig.~\ref{fig:sq}(c), the 5x5 grid is
clearly visible in the maximal value map, though the localisation of each
connection density map is still not sufficient for the contour map to emerge.
%
The contour map emerges by 0.81 s and is well formed by 1.77 s
[Figs.~\ref{fig:sq}(d) \& (e)]. Fig.~\ref{fig:sq}(e) shows the interaction
parameters for each TC type ($\gamma$). The values of $\gamma$ were chosen
with the intention of forming a grid; the interactions for $i =$ 1, 5, 21 \&
25 (the corners) were strongest and the interaction parameters for $i=$13 (the
centre) were zero. Fig.~\ref{fig:sq}(d) shows the mapping between colour and
TC type $i$ and can be used, in conjunction with Fig.~\ref{fig:sq}(e), to see
that $\gamma_{1,1}=-2$, $\gamma_{1,2}=-2$, $\gamma_{3,1}=0$,
$\gamma_{3,2}=-2$, and so on.

The simulation shown in Fig.~\ref{fig:sq} demonstrates that complex structure
can form in a system with just two guidance gradients \cmnt{(provided that
  interactions with each gradient occur in both senses; positive and negative)}.

@@

\subsubsection*{Stronger branching competition}

@@

\begin{figure*}
\begin{center}
\includegraphics[width=0.7\textwidth]{../boundaries/barreloids_haidarliu_graph.png}
\end{center}
%\vspace{-0em}
\caption{Barreloid centres taken from \cite{haidarliu_size_2001} are shown as
  black circles. It is assumed that there are two linearly increasing,
  orthogonal signalling gradients (molecules A \& B) in the thalamus (side
  graphs), such that the location of each barreloid within the graded thalamic
  signal field can differentiate the efferent axons, with the result that
  axons from each barreloid interact to varying degrees with cortical
  signalling gradients. Coloured arrows show, for each barreloid, the
  magnitude (and sign) of the interaction that its thalamocortical projection
  will have with the signalling gradients in the cortex. The mappings from
  barreloid location to interaction strength are linear.}
\label{fig:barreloids}
\end{figure*}

\begin{figure*}
\begin{center}
\includegraphics[width=\textwidth]{./realistic_barrels.png}
\end{center}
%\vspace{-0em}
\caption{Realistic barrel pattern}
\label{fig:barrels}
\end{figure*}

@@

Linear, circ linear, exponential, etc.

\subsection*{The barrel field reacts to simulated whisker plucking}

A section on whisker plucking?

%\subsubsection*{With surrounding large domains}
%Maybe, put this in to remind myself that I did this. It looks nice.

%\subsection*{Plucking experiments}
%Include results of plucking a single whisker, a row of whiskers and a
%column of whiskers.

%\subsection*{Any parameter sweeps necessary to show behaviour is
%  robust to changes in params}

\subsection*{Extending the model to capture row and arc development}

\subsubsection*{Manipulating Competition: Group competition stronger/weaker than inter-group competition}

Just an idea - have two separate epsilons.

\subsubsection*{Manipulating Guidance: Relative interaction strengths}

One possibility is to modify the strength of the guidance interactions of the
two, orthogonal guidance gradients. In the simulation we modify the magnitude
of the expression of each guidance molecule, although this is equivalent to,
and could stand for a more or less vigourous movement of axons in response to
un-modified guidance gradients.

\subsubsection*{Manipulating Guidance: Delayed onset in one axis}

Another idea; begin with only one guidance gradient, or one strong and
one very weak guidance gradient, and begin the development in this
form, so as to encourage the formation of rows. Subsequently, turn on
the second strong guidance signal.

%\begin{equation}
%\chi=\frac{\epsilon_i  a_i}{N-1} \sum_{j \ne i}^{N} a_j(\mb{x}, t)^l
%\end{equation}

%And normalization step

%\begin{equation}\label{eq:norm_comp8}
%a_i' = n_e \; \frac {a_i} {\textstyle \sum_j^{n_e} a_{i,j} }
%\end{equation}

%($a_i'$ becomes $a_i$ in the next computational step.)

@@

\bibliography{BarrelEmerge}

@@

%%%%% ORIGINAL EQUATIONS MAINTAINED AS NOTES HERE FOR REFERENCE

\section*{App 1: Karbowski system with N TCs and M gradients}
%
\begin{equation} \label{eq:Karb2D_dc}
\frac{\partial c_i(\mb{x},t)}{\partial t} = -\alpha c_i(\mb{x},t) + \beta n(\mb{x},t)
[a_i(\mb{x},t)]^k
\end{equation}
%
\begin{equation} \label{eq:Karb2D_conserve}
n(\mb{x},t) = 1 - \sum_{i=1}^{N} c_i(\mb{x}, t)
\end{equation}
%
\begin{equation} \label{eq:Karb2D_da}
\frac{\partial a_i(\mb{x},t)}{\partial t}
= \nabla\vcdot\mb{J}_i(\mb{x},t) - \frac{\partial c_i(\mb{x},t)}{\partial t}
\end{equation}

with the flux current

\begin{equation} \label{eq:Karb2D_J}
\mb{J}_i(\mb{x},t) = D \nabla a_i(\mb{x},t) - a_i
\sum_{j=1}^M \big(\gamma_{i,j} \nabla\rho_j(\mb{x}) \big)
\end{equation}

The boundary condition applied to these equations is:

\begin{equation}
\mb{J}_i(\mb{x},t) \bigg\rvert_{boundary} = 0
\end{equation}

\section{App 2: Karbowski system with N TCs and M gradients plus competition}

%
\begin{equation} \label{eq:dc2}
\frac{\partial c_i(\mb{x},t)}{\partial t} = -\alpha c_i(\mb{x},t) + \beta n(\mb{x},t)
[a_i(\mb{x},t)]^k
\end{equation}
%
\begin{equation} \label{eq:conserve}
n(\mb{x},t) = 1 - \sum_{i=1}^{N} c_i(\mb{x}, t)
\end{equation}
%
\begin{equation} \label{eq:da2}
\frac{\partial a_i(\mb{x},t)}{\partial t}
= \nabla\vcdot\mb{J}_i(\mb{x},t) - \frac{\partial c_i(\mb{x},t)}{\partial t}- \frac{\epsilon_i  a_i}{N-1} \sum_{j \ne i}^{N} a_j(\mb{x}, t)^l
\end{equation}

And normalization step

\begin{equation}\label{eq:norm_comp8}
a_i' = n_e \; \frac {a_i} {\textstyle \sum_j^{n_e} a_{i,j} }
\end{equation}

($a_i'$ becomes $a_i$ in the next computational step.)

>>>>

% Brief report style paper

% It's about how a somatotopically ordered pattern can be
% re-established in the cortex based on information carried with
% thalamocortical axons afferent from the ventrobasal thalamus, but
% without specifying absolute positions as centre-points for the
% barrels. Everything in the paper should be tailored towards arguing
% this idea.

\documentclass[9pt,twocolumn,twoside,lineno]{pnas-new}
\templatetype{pnasresearcharticle}

\title{Self-organization of whisker barrels in simulation requires competition between thalamocortical axons}
% While writing, don't stop for errors
%\nonstopmode

% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
\author[a,1]{Sebastian~S.~James}
\author[a]{Stuart~P.~Wilson}

\affil[a]{Department of Psychology, The University of Sheffield, Sheffield, United Kingdom.}

% $^{*}$Corresponding author email: Seb.James@Sheffield.ac.uk\vspace{1em}\\

% Please give the surname of the lead author for the running footer
\leadauthor{James}

% Please add a significance statement to explain the relevance of your work
\significancestatement{The mystacial whiskers of the rat are a complex, ordered set of structures whose pattern is repeated several times in separate locations in the brain. We contribute a mechanism by which their spatially ordered pattern can be recreated in a self-organising process which requires minimal genetic pre-determination.}

\correspondingauthor{\textsuperscript{1}To whom correspondence should be addressed. E-mail: seb.james\@sheffield.ac.uk}

% At least three keywords are required at submission. Please provide three to five keywords, separated by the pipe symbol.
\keywords{Keyword 1 $|$ Keyword 2 $|$ Keyword 3 $|$ ...}

\begin{abstract}
Understanding how genetic information and spontaneous pattern formation
together give rise to the whisker barrels is an important challenge for
developmental neurobiology. The barrel patterning that coalesces in layer IV
has been described formally as a Voronoi tesselation, which is indicative of
an underlying developmental process based on local competitive
interactions. If pattern formation relies only on such intrinsic
self-organising processes, how can the somatotopic order at the periphery be
maintained in the projections from brainstem to thalamus and from thalamus to
cortex? To investigate, we began by extending a computational model of
neocortical arealization by Karbowski \& Ermentrout (2004; J. Comp. Neurosci.)
to generate fields on a two-dimensional cortical sheet. Central to this model
is an assumption that axonal branching and synaptogenesis are coupled, and
that thalamic projections compete for cortical territory under the influence
of molecular signalling gradients. We found that these processes alone are
incapable of generating patterns of cortical fields that are more intricate
than the patterns already present in molecular guidance signals. However, a
full barrel-field pattern emerges under the guidance of as few as four
signalling gradients when the model is extended to introduce competition
between the branches of axons that originate from different thalamic
barreloids. We show that the key requirements for the emergence of realistic
(Voronoi tesselation) patterning are i) at each cortical location
thalamocortical projections compete for a limited number of available
connections, ii) at each location the axonal branching rate of each
thalamocortical projection is reduced by the total branching density of the
other projections, and iii) the density of axon branching for each
thalamocortical projection is conserved. We demonstrate how a purely
self-organising system can initialise a pattern which can then be faithfully
replicated in multiple neural structures. We show that this minimal model of
barrel self-organisation provides a parsimonious account of data from a range
of classic studies in which intrinsic and extrinsic factors have been
experimentally manipulated.
\end{abstract}

\dates{This manuscript was compiled on \today}
\doi{\url{www.pnas.org/cgi/doi/10.1073/pnas.XXXXXXXXXX}}

\begin{document}

% Blue for comments
\newcommand{\cmnt}[1]{\textcolor{blue}{#1}}
% Divergence symbol (Del dot Del)
\newcommand{\dvrg}{\nabla\vcdot\nabla}
% Emphasis shortcut
\newcommand{\e}{\emph}
% Bold shortcut
\newcommand{\bol}{\textbf}
% Math bold
\newcommand{\mb}[1]{\mathbf{#1}}
\makeatletter
\newcommand*\vcdot{\mathpalette\vcdot@{.35}}
\newcommand*\vcdot@[2]{\mathbin{\vcenter{\hbox{\scalebox{#2}{$\m@th#1\bullet$}}}}}
\makeatother

\maketitle
\thispagestyle{firststyle}
\ifthenelse{\boolean{shortarticle}}{\ifthenelse{\boolean{singlecolumn}}{\abscontentformatted}{\abscontent}}{}

@@

\dropcap{S}patial patterns in neural connectivity provide clues about
the constraints under which brains evolve and develop. Whether or not
patterning governs information processing, models of the development
of information processing circuits can be validated in terms of their
ability to account for biological patterning
(\citealp{purves_iterated_1992}; see also
\citealp{wilson_what_2015,bednar_cortical_2016}). One of the most
distinctive patterns of neural connectivity is found in the rodent
barrel cortex
(\citealp{woolsey_pattern_1948,woolsey_structural_1970,welker_structure_1974},
see \citealp{fox_barrel_2008}). Understanding how this particular
pattern forms might shed light on the interplay of genetic and
extrinsic factors that shape cortical circuits, and on a wealth of
anatomical, physiological, and behavioural data obtained using this
model system.

@@

We first verified that the system given by Eqs.~\ref{eq:dc} \& \ref{eq:da}
described a two dimensional extension of the axon branching model in
\citet{karbowski_model_2004}. We showed that in the presence of three guidance
gradients varying only in one dimension (along the long axis of an elliptical
domain), five thalamocortical axon types, each interacting uniquely with the
three guidance gradients, would segregate into five separate regions. We
investigated a number of configurations of the guidance gradients, varying
their shape, $\rho(\mb{x})$, and number, $M$, and their effect on a varying
number of thalamocortical projections, $N$. In general, we were not able to
produce a complexity of structure in the resulting patterns of the
connections, $c(\mb{x},t\rightarrow\infty)$, which was not present in the
guidance patterns.

@@

To introduce competition between the \emph{branching} of thalamocortical axon
types, we introduced the new term $\chi$ given by
%
\begin{equation} \label{eq:branch_comp}
\chi(\mb{x}, t) = - \frac{\epsilon  a_i(\mb{x}, t)}{N-1} \sum_{i' \ne i}^{N} a_{i'}(\mb{x}, t).
\end{equation}
%
which leads to a reduction in the rate of increase of branching density for
axon type $i$ in regions where the branching density of \emph{other} axon types
is high.

@@

With the addition of the competition term, $\chi$, it is necessary to
introduce a normalisation step, in order to prevent one TC type from
{out-competing every other TC type. Normalisation represents the idea
that for a fixed number of afferent thalamocortical axons, the overall
number of branches must have some upper limit; given the finite
resources available in each axon for building new processes. We
divisively normalize the branching density, dividing it at each
computational step by the mean branching density per unit area. It is
then multiplied by the branching density per unit area at time 0, to
account for the possibility that there might be more axons afferent
for TC type $1$ cf.~TC type $2$:
%
\begin{equation} \label{eq:norm}
  a_i(\mathbf{x}, t) = \int_\Omega  a_i(\mathbf{x}, 0) d\sigma \; \frac {a'_i(\mathbf{x}, t)} {\int_\Omega
  a'_i(\mathbf{x}, t) d\sigma}
\end{equation}
%
The prime symbols indicate the use of the intermediate value of $a_i$
after it has been computed according to Eqs.~\ref{eq:dc}--\ref{eq:branch_comp}
on the basis of the system state at time $t-{\delta}t$.

@@

@@

\begin{figure*}
\begin{center}
\includegraphics[width=0.7\textwidth]{../../boundaries/barreloids_haidarliu_graph.png}
\end{center}
%\vspace{-0em}
\caption{Barreloid centres taken from \cite{haidarliu_size_2001} are shown as
  black circles. It is assumed that there are two linearly increasing,
  orthogonal signalling gradients (molecules A \& B) in the thalamus (side
  graphs), such that the location of each barreloid within the graded thalamic
  signal field can differentiate the efferent axons, with the result that
  axons from each barreloid interact to varying degrees with cortical
  signalling gradients. Coloured arrows show, for each barreloid, the
  magnitude (and sign) of the interaction that its thalamocortical projection
  will have with the signalling gradients in the cortex. The mappings from
  barreloid location to interaction strength are linear.}
\label{fig:barreloids}
\end{figure*}

\begin{figure*}
\begin{center}
\includegraphics[width=\textwidth]{../longpaper/realistic_barrels.png}
\end{center}
%\vspace{-0em}
\caption{Realistic barrel pattern}
\label{fig:barrels}
\end{figure*}

@@

@@

\bibliography{../BarrelEmerge}

@@
